---
title: 'Exercise: Make this document'
author: "YOUR NAME"
date: "TODAY'S DATE"
output: pdf_document
---

***

Let's see if you can make a document that looks *exactly like this* using **R Markdown.** Try to copy this document verbatum. Set the output to "pdf_document"

***

###1. Insert this plot from the `iris` dataset

Here, I want you to insert a plot of three species of iris from the `iris` dataset. I have suppressed the printing of the code here. 
Some clues:

* It's easy to produce boxplots using the formula syntax inside the `plot()` function.
* Use the code chunk option that prevents the code from being printed
* Use the code chunk option that shrinks the plot by 50%
* The three shades I used are "gray100", "gray75" and "gray50".


```{r, echo=FALSE, out.width="50%", fig.align="center"}
plot(Sepal.Length~Species, data=iris, col=c("gray100", "gray75", "gray50"), main="Sepal Length of Three Species of Iris")
```


###2. Conduct and report the results of an analysis
We can conduct a linear regression between the speed of a car and the distance it takes for the car to stop. This data is available in a dataset called `cars` that is pre-installed in R. 

```{r}
head(cars)
```

Let's plot it:
```{r, out.width="50%", fig.align="center"}
plot(dist~speed, data=cars, pch=19)
```

We can conduct the linear regression with *speed* as the independent variable and *distance* as the dependent variable:

```{r}
fit.cars=lm(dist~speed, data=cars)
summary(fit.cars)
```


###3. Create a table to present results
Let's organize the results as a table:


Slope | *F* (d.f.) | Adjusted *R^2^* | *P* 
----- | -----------| ----------------| ----
3.93  |89.6 (1,48) |   0.64          | <0.001


There are other ways to make pretty tables, but we'll just go with the simple markdown format presented in the Reference Guide. 
